# Lab 8: One-way ANOVA

<div class="marginnote">
This lab is modified and extended from [Open Stats Labs](https://sites.trinity.edu/osl). Thanks to Open Stats Labs (Dr. Kevin P. McIntyre) for their fantastic work.
</div>

## How to not think about bad memories by playing Tetris

This lab activity uses the open data from Experiment 2 of James et al. (2015) to teach one-way ANOVA with planned comparisons. Results of the activity provided below should exactly reproduce the results described in the paper.

### STUDY DESCRIPTION

Following traumatic experiences, some people have flashbacks, which are also called "intrusive memories"" and are characterized by involuntary images of aspects of the traumatic event. Although people often try to simply forget traumatic memories, this approach is not very effective. Instead, previous research suggests that a better approach may be to try to change aspects of the memory after it is formed. For example, some research shows that traumatic memories can be altered and weakened to the point that they are no longer intrusive.

Because intrusive memories of trauma are often visual in nature, James and colleagues (2015) sought to explore whether completing a visuospatial task (e.g., tetris) after a memory was formed would interfere with the storage of that memory, and thereby reduce the frequency of subsequent intrusions. They hypothesized that  only participants who complete a visuospatial task after reactivation of the traumatic memories would experience a  reduction in intrusive memories. In comparison, simply completing a visuospatial task (without reactivation) or reactivation (without a visuospatial task), would not reduce the occurrence intrusive memories.

In other words, if you play tetris shortly after you were remembering bad memories, playing tetris might weaken those memories, which could cause you experience those kinds of intrusive memories less often in the future.

### Study Methods

To test their hypothesis, the  autho rs conducted an experiment ( N = 72,  n = 18 per condition). The procedure is summarized as follows:

**Trauma Film**: All participants viewed a series of video clips of graphic violence (e.g., a person getting hit by a van while using his phone as he crosses the road) as a way to create memories that should become intrusive memories. Participants then went home and recorded the number of intrusive memories they experienced over the next 24 hours. Because this is before the experimental manipulations, all groups were predicted to have an  equal occurrence of intrusive memories during the first 24-hours (called Day 0).

**Experimental Task**: After this 24-hour period, the participants returned to the lab and completed the experimental task. The experimenters randomly assigned participants to ONE of the following conditions:

1. No-task control: These participants completed a 10-minute music filler task.
2. Reactivation + Tetris: These participants were shown a series of images from the trauma film to reactivate the traumatic memories (i.e., reactivation task). After a 10-minute music filler task, participants played the video game Tetris for 12 minutes.
3. Tetris Only: These participants played Tetris for 12 minutes, but did not complete the reactivation task.
4. Reactivation Only: These participants completed the reactivation task, but did not play Tetris.

**Intrusive Memories**: All participants were asked to record the number of intrusive memories that they experienced over the next seven days (Days 1 to 7).

After the seven days had passed, participants completed an Intrusion-Provocation Task, in which they were shown blurred images from the trauma film and asked to indicate whether the blurred image triggered an intrusive memory.


## Lab Skills Learned

## Important Stuff
- citation: James, E. L., Bonsall, M. B., Hoppitt, L., Tunbridge, E. M., Geddes, J. R., Milton, A. L., & Holmes, E. A. (2015). Computer game play reduces intrusive memories of experimental trauma via reconsolidation-update mechanisms. Psychological Science, 26, 1201-1215.
- [Link to .pdf of article](http://journals.sagepub.com/stoken/default+domain/hQ2W4fbPrZVJ7eyNJaqu/full)
- [Data in .csv format](https://drive.google.com/open?id=0Bz-rhZ21ShvOM1cxWUpUNlQ0UlE)
- [Data in SPSS format](https://drive.google.com/file/d/0Bz-rhZ21ShvOZ1lvQ0dQekZGWU0/view?usp=sharing)

## R

### Load the data

Remember that any line with a \# makes a comment and the code does not run. Below is how to load the .csv data from the online repository, or from a local file (you need to change the file path to where the local file is, if you downloaded it). The data contains all of the measures and conditions from Experiment 2 in the paper.

```{r}
library(data.table)
#fread("https://raw.githubusercontent.com/CrumpLab/statisticsLab/master/data/Jamesetal2015Experiment2.csv")
all_data <- fread("data/Jamesetal2015Experiment2.csv")
```

### Inspect the dataframe

This will give you a big picture of the data frame. Click the button to view it in your browser, then take a look to see what is in it. 

```{r, eval=F}
library(summarytools)
view(dfSummary(all_data))
```

### Get the data you need

Again we have only the data from Experiment 2, so we don't need to get rid of any rows of the data. But we do need look at the columns to see how the independent variable and dependent variables were coded. 

#### The independent variable

There was one important independent variable, and it had four levels. The first column in the data frame is called condition, and it has four levels. The levels are 1, 2, 3, and 4 in the dataframe. These will correspond to the four levels in shown in figure: 

- No-task control
- Reactivation Plus tetris
- Tetris only
- Reactivation only

Each of these refer to what subjects did after they watched the traumatic film. But, which of these correspond to the numbers 1 to 4 in the dataframe? It turns out the are in the right order, and 1 to 4 refer to:

1. No-task control
2. Reactivation Plus tetris
3. Tetris only
4. Reactivation only

Let's do ourselves a favor and rename the levels of the `Condition` column with words so we know what they refer to. First, convert the `Condition` column to a factor, then rename the levels. Then.


```{r}
all_data$Condition <- as.factor(all_data$Condition)
levels(all_data$Condition) <- c("Control",
                                "Reactivation+Tetris", 
                                "Tetris_only",
                                "Reactivation_only")
```

#### The dependent variable

The authors showed two figures, one where they analysed


## Excel

How to do it in Excel

## SPSS

How to do it in SPSS

## Matlab

How to do it in Matlab


